---
title: "Power Analysis for Elena's K Proposal"
output: html_document
author: Christof Seiler
date: August, 2017
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

Install packages.

```{r warning=FALSE, message=FALSE}
pkgs_needed = c("ggplot2","reshape2","tibble","dplyr","magrittr","stringr","devtools",
                "parallel")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  source("http://bioconductor.org/biocLite.R")
  biocLite(letsinstall)
}
```

Load packages.

```{r load_packages}
library("ggplot2")
library("reshape2")
library("tibble")
library("dplyr")
library("magrittr")
library("stringr")
library("devtools")
library("parallel")
```

## Power Simulations

* Total number of makers: 36
* Activating markers: 10 to 15
* Number of NK cells: 5,000 to 10,000
* Fold change: 1.5 (e.g. the mean intensity count is 100 in condition A and 150 in condition B/A=1.5)

### Study 1 (Aim 1)

* 3 conditions: 
    * longterm nonprogressors (n=18)
    * typical progressors (n=61)
    * treated patients (n=24)
* 3 to 4 timepoints per patient

```{r study_1}
```

### Study 2 (Aim 2)

* 2 conditions:
    * before treatment (n=24)
    * after treatment (n=24)

```{r study_2}
```

### Study 3 (Aim 2)

* 50 patients, on treatment
* 3 timepoints per patient

```{r study_3}
```

## Session Info

```{r session_info}
session_info()
```
