---
title: "Randomization"
output: html_document
author: Christof Seiler
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

Reverse engineer how CyTOF converts integer counts into real numbers.

## Prepare Packages

Install packages.

```{r warning=FALSE, message=FALSE}
pkgs_needed = c("ggplot2","reshape2","tibble","dplyr","magrittr",
                "devtools","flowCore")
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  source("http://bioconductor.org/biocLite.R")
  biocLite(letsinstall)
}
```

Load packages.

```{r load_packages, warning=FALSE, message=FALSE}
library("ggplot2")
library("reshape2")
library("tibble")
library("dplyr")
library("magrittr")
library("devtools")
library("flowCore")
theme_set(theme_bw())
```

## Background

After personal conversation with Mike, I think this is what is going on during randomization.

1. If count is zero, no randomization
2. If count is non-zero, draw number between 0 and 1 (excluding 0 and 1) and subtract it from its observed count

So what we have to do is to round up all non-zero counts to the next higher integer. This will perfectly reconstruct the original values.

## Try Derandomize Counts

Load fcs files before and after randomization. It looks like only the first 25 cells are the same in before and after fcs files. Use `ceiling` to derandomize and check if it works.

```{r load_fcs}
before = "EV02_4a_1_00_norandom_cells_found.fcs"
after = "EV02_4a_1_00.FCS"
marker = "Cd111Di"
fcs_before = read.FCS(before,transformation = FALSE)
colnames(fcs_before)
fcs_after = read.FCS(after,transformation = FALSE)
colnames(fcs_after)
exprs_before = fcs_before@exprs[,marker]
length(exprs_before)
exprs_after = fcs_after@exprs[,marker]
length(exprs_after)
expr_tb = tibble(before = head(exprs_before,n = 25),
                 after = head(exprs_after,n = 25))
expr_tb %<>% mutate(diff = after - before)
expr_tb %<>% mutate(derandomized = ceiling(after))
expr_tb %<>% mutate(diff_derandomized = derandomized - before)
expr_tb %>% print(n = Inf)
```

```{r session_info}
session_info()
```
